State of the Practice for Lattice Boltzmann Method Software

### 2022-04-22 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Edits up to, but not including, Section 2.5

We developed a methodology for evaluating the state of the practice of research
software \citep{SmithEtAl2021, SmithAndMichalski2022}. This methodology can be
applied to a specific domain of research software, which in the current case is
LBM software.  Since we group best practices around conventional software
qualities, the first section below provides definitions for the qualities of
interest (Section~\ref{softwarequalities}), along with information on how to
measure these qualities. The following section (Section~\ref{Sec_OverallProcess})
provides an overview of the steps we used to select, measure and compare LBM
software.  The remaining sections provide details on each step in the
methodology.

@@

To ensure our assessment of best practices is organized and complete, our
measurements are centred around software qualities.  For completeness we used
the the list of software engineering qualities from \citet[p.\
17--33]{GhezziEtAl2003}, excluding interoperability, productivity and
timeliness.  We excluded interoperability because we are focusing on LBM
software alone, not its interaction with other software.  The other two
qualities were excluded because they are process qualities, and we are measuring
product qualities.  To the list from \citet{GhezziEtAl2003} we added
installability, since it is so critical to the success of any practical project,
and reproducibility, since it is so important for research software. Below we
provide information on each quality of interest, including information on how
they are measured.

@@

	\item[Installability] Installability is measured by the effort required for
	the installation, uninstallation or reinstallation of a software product in
	a specified environment \citep{ISO/IEC25010, lenhard2013measuring}. In our
	case, effort includes the time spent finding and understanding the
	installation instructions, the person-time and resources spent performing
	the installation procedure, and the absence or ease of overcoming system
	compatibility issues. An installability score improves if the software has a
	means to validate the installation and instructions for uninstallation.

	\item[Correctness] A software program is correct if it behaves according to
	its stated specifications \citep[p.\ 17]{GhezziEtAl2003}. This requires that
	the specification is available. Research software is unlikely to have an
	explicit specification, since this is not common practice in the field. As a
	consequence, the correctness of software often cannot be measured directly.
	We assess correctness indirectly by looking for the following: availability
	of a requirements specification, reference to domain theory, and explicit
	use of tools and/or techniques for building confidence, such as
	documentation generators and software analysis tools.

	\item[Verifiability] An artifact (like code, or a theory manual) is
	verifiable if, and only if, every statement therein is verifiable. A
	statement is verifiable if, and only if, there exists some finite
	cost-effective process with which a person or machine can check that each
	statement is correct (adapted from recommended practice for requirements
	specifications \citep{IEEE1998}). Similarly to correctness, verifiability is
	correlated with the availability of a specification and with reference to
	domain knowledge. A good measure of verifiability is further correlated with
	the availability of well written tutorials that include expected output,
	with software unit testing, and with evidence of Continuous Integration
	(CI). In our process we measure correctness and verifiability together.

@@

	\item[Performance] Performance is measured by the degree to which a system
	or component accomplishes its designated functions within given constraints,
	such as speed (database response times, for instance), throughput
	(transactions per second), capacity (concurrent usage loads), and timing
	(hard real-time demands) \citep{IEEEStdGlossarySET1990, wiegers2003softreq}.
	In this state of the practice assessment performance was not directly
	quantitatively measured. Instead the documentation of each software package
	was observed for information alluding to consideration of performance, such
	testing results, or parallelization instructions.

@@

	\item[Maintainability] A measure of maintainability is the effort with which
	a software system or component can be modified to correct faults, improve
	performance or other attributes, and satisfy new requirements
	\citep{IEEEStdGlossarySET1990, boehm2007software}. In the current work,
	maintainability is measured by the quality of the documentation artifacts,
	and the presence of version control and issue tracking. These artifacts can
	greatly decrease the effort needed to modify software. There are many
	documentation artifacts that can improve maintainability, including user and
	developer manuals, specifications, README files, change logs, release notes,
	publications, forums, and instructional websites. We also asked developers
	if they considered the ease of future changes when developing the software
	packages.

	\item[Reusability] Reusability refers to the extent to which components of a
	software package can be used with or without adaptation in other software
	packages \citep{kalagiakos2003non}. A good indicator of reusability is a
	large number of easily reusable components. Therefore, we look for high
	modularization, which is defined as the presence of smaller components with
	well defined interfaces. For this state of the practice assessment, we
	assume that a good measure of reusability correlates with a high
	number of code files, and the availability of API documentation.

	\item[Understandability] Understandability is measured by the capability of
	the software package to enable the user to understand its suitability and
	function \citep{ISO9126}. It is an artifact-dependent quality.
	Understandability is different for the user-interface, source code, and
	documentation. In this state of the practice analysis, understandability
	focuses on the source code. It is measured by the consistency of a
	formatting style, the extend of modularization, the explicit identification
	of coding standards, the presence of meaningful identifiers, and the clarity
	of comments.

	\item[Visibility and Transparency] Visibility and transparency refer to the
	extent to which all of the steps of a software development process, and the
	current status of it, are conveyed clearly \citep[p.\ 32]{GhezziEtAl2003}.
	In this state of the practice assessment a good measure of visibility and
	transparency correlates with a well defined development process,
	documentation of the development process and environment, and software
	version release notes.

	\item[Reproducibility] Software achieves reproducibility if another
	developer can take the original code and input data, run it, and obtain the
	original output \citep{BenureauAndRougier2017}. We measured reproducibility
	qualitatively by asking developers if they have any concern that their
	computational results won't be reproducible, and if they have taken any
	steps to ensure reproducibility.

@@

To answer~\rqref{RQ_WhatProjects} we needed to identify existing LBM projects.
The candidate software was found through search engine queries targeting
authoritative lists of software. We found LBM software listed on
\href{https://github.com/} {GitHub} and \href{https://swmath.org/} {swMATH}, as
well as through scholarly articles. The Domain Expert
(Section~\ref{sec_vet_software_list}) was also engaged in selecting the
candidate software.

The following properties were considered when creating the list and reviewing
the candidate software:

@@

The initial list had 45 packages, including a few that were later found to not
have publicly available source code, or to be incomplete projects.

@@

To reduce the length of the list of software packages and to
answer~\rqref{RQ_WhatProjects}, we filter the initial list of 45 packages.  The
following filters were applied in the priority order listed.

\begin{enumerate}
	\item Scope: Software is removed by narrowing what functionality is
	considered to be within the scope of the domain.
	\item Usage: Software packages are eliminated if their installation
	procedure was missing or not clear and easy to follow.
	\item Age: The older software packages (age being measured by the last date
	when a change was made) are eliminated, except in the cases where an older
	software package appears to be highly recommended and currently in use. 
\end{enumerate}

For the third item in the above filter, software packages were characterized as
`alive' if their related documentation had been updated within the last 18
months. Packages were categorized as `dead' if the last update was more than 18
month ago.

@@

Filtering by scope, usage, and age decreased the size of the list to 23
packages. Many of the 22 packages that were removed could not be tested as there
was no installation guide, they were incomplete, source code was not publicly
available, a license was needed, or the project was out of scope or not up to a
standard that would support incorporating them into this study. These eliminated
software packages are listed in the Appendix of \citet{Michalski2021}. Of the
remaining 23 packages that were studied, some were kept on the list despite
being marked as dead due to their prevalence on authoritative lists on LBM
software and due to their surface excellence.  After the initial measurement
exercise, which took place in mid 2020, we added one more package in January
2022: Musubi.  Musubi did not come up in our initial search, but once we learned
of it, we felt that this report was incomplete without including it.  With the
addition of Musubi we did not update the manual collection of data for the
previous packages, due to time constraints, but we did redo all data collection
that was automated by tools.

The final list of 24 software packages that were analyzed in this project can be
found in two tables. Table~\ref{alivepackages} lists packages that fell into the
`alive' category as of mid 2020 (January 2022 for Musubi), and
Table~\ref{deadpackages} lists packages that were `dead' at that time. The
tables include a hyperlink to each project and, when available, citations for
relevant publications.  The final list of software packages shows considerable
variation in purpose, size, user interfaces, and programming languages. For
example, the OpenLB software package is predominantly a \CC{} package that makes
use of hybrid parallelization and was designed to address a range of CFD
problems \citep{heuveline2009towards}. The software package pyLBM is an
all-in-one Python language package for numerical simulations
\citep{graille2017pylbm}. ESPResSo is an extensible simulation package that is
specifically for research on soft matter, and is written in \CC{} and Python
\citep{weik2019espresso}. The HemeLB package is used for efficient simulation of
fluid flow in several medical domains, and is written predominantly in C, \CC,
and Python \citep{mazzeo2008hemelb}.

@@

The top four packages, ESPREsSo, Ludwig, Palabos, and OpenLB, have all of the
common artifacts, except Palabos does not have theory notes and only three
(ESPREsSo, Ludwig, Palabos) of the packages appear to use a version control
system.  The top four packages also have most of the uncommon artifacts. At
the time of data collection, only one of the four packages (Palabos) did not
have a user manual or guide, but there was a broken link on the package website
indicating that such an artifact might exist. This broken link was later fixed,
but this is not reflected in our data because it was not present at the time of
data collection (mid 2020). Despite the broken link, Palabos does have a
detailed and informative website. As far as we could observe, Ludwig does not
appear to have a publicly visible design and OpenLB does not appear to use a
version control system. However, OpenLB's use of version numbers suggests that
version control may be used privately by the developers.

>>>>

We developed a methodology for evaluating the state of the practice of research
software \citep{SmithEtAl2021, SmithAndMichalski2022} and applied it to the
specific domain of LBM software.  Since we group best practices around
conventional software qualities, the first section below provides definitions
for the qualities of interest (Section~\ref{softwarequalities}), along with
information on how to measure these qualities. The following section
(Section~\ref{Sec_OverallProcess}) provides an overview of the steps we used to
select, measure and compare LBM software.  The remaining sections provide
details on each step in the methodology.

@@

To ensure our assessment of best practices is organized and complete, we centre
our measurements around software qualities. For completeness, we use the list of
software engineering qualities from \citet[p.\ 17--33]{GhezziEtAl2003},
excluding interoperability, productivity and timeliness. We excluded
interoperability because our focus is on LBM software alone, not its interaction
with other software. Our exclusion of the other two qualities is because they
are process qualities, and our focus is on product qualities. We added
installability to our list since potential users will ignore software they
cannot install.  Since the scientific method depends on reproducibility, we also
added this quality. Below we provide information on each quality, including
details on how they are measured.

@@

	\item[Installability] Installability is measured by the effort required for
	the installation, uninstallation or reinstallation of a software product in
	a specified environment \citep{ISO/IEC25010, lenhard2013measuring}. In our
	case, the effort includes the following: time spent finding and
	understanding the installation instructions, the person-time and resources
	spent performing the installation procedure, and the absence or ease of
	overcoming system compatibility issues. An installability score improves if
	the software has means to validate the installation and instructions for
	uninstallation.

	\item[Correctness] A software program is correct if it behaves according to
	its stated specifications \citep[p.\ 17]{GhezziEtAl2003}. Therefore,
	measuring correctness requires that a specification is available. However,
	research software is unlikely to have an explicit specification. As a
	consequence, correctness often cannot be measured directly. We assess
	correctness indirectly by looking for the following: availability of a
	requirements specification, reference to domain theory, and explicit use of
	tools and/or techniques for building confidence, such as documentation
	generators and software analysis tools.

	\item[Verifiability] An artifact (like code, or a theory manual) is
	verifiable if, and only if, every statement therein is verifiable. A
	statement is verifiable if, and only if, there exists some finite
	cost-effective process with which a person or machine can check that each
	statement is correct (adapted from recommended practice for requirements
	specifications \citep{IEEE1998}). Similarly to correctness, verifiability is
	correlated with the availability of a specification and with reference to
	domain knowledge. A good measure of verifiability is further correlated with
	the following: availability of well-written tutorials that include expected
	output, software unit tests, and evidence of Continuous Integration (CI).
	Our process measures correctness and verifiability together.

@@

	\item[Performance] Performance is measured by the degree to which a system
	or component accomplishes its designated functions within given constraints,
	such as speed (database response times, for instance), throughput
	(transactions per second), capacity (concurrent usage loads), and timing
	(hard real-time demands) \citep{IEEEStdGlossarySET1990, wiegers2003softreq}.
	In this state of the practice assessment performance is not directly
	quantitatively measured. Instead we check the documentation for each
	software package looking for information alluding to consideration of
	performance, such as testing results or parallelization instructions.

@@

	\item[Maintainability] A measure of maintainability is the effort with which
	a software system or component can be modified to correct faults, improve
	performance or other attributes, and satisfy new requirements
	\citep{IEEEStdGlossarySET1990, boehm2007software}. In the current work, we
	measure maintainability by the quality of the documentation and the presence
	of version control and issue tracking. There are many documentation
	artifacts that can improve maintainability, including user and developer
	manuals, specifications, README files, change logs, release notes,
	publications, forums, and instructional websites. We also asked developers
	if they considered the ease of future changes when developing the software
	packages.

	\item[Reusability] Reusability refers to the extent to which components of a
	software package can be used with or without adaptation in other software
	packages \citep{kalagiakos2003non}. A good indicator of reusability is a
	large number of easily reusable components. Therefore, we look for high
	modularization, which we define as the presence of small components with
	well defined interfaces. For this state of the practice assessment, we
	assume that a good measure of reusability correlates with a high number of
	code files, and the availability of API documentation.

	\item[Understandability] Understandability is measured by the capability of
	the software package to enable the user to understand its suitability and
	function \citep{ISO9126}. It is an artifact-dependent quality.
	Understandability is different for the user interface, source code, and
	documentation. In this state of the practice analysis, understandability
	focuses on the source code. We  measure it by the consistency of  formatting
	style, the extent of modularization, the explicit identification of coding
	standards, the presence of meaningful identifiers, and the clarity of
	comments.

	\item[Visibility and Transparency] Visibility and transparency refer to the
	extent to which all of the steps of a software development process, and its
	current status, are conveyed clearly \citep[p.\ 32]{GhezziEtAl2003}. In this
	state of the practice assessment a good measure of visibility and
	transparency correlates with a well-defined development process,
	documentation of the development process and environment, and software
	version release notes.

	\item[Reproducibility] Software achieves reproducibility if another
	developer can take the original code and input data, run it, and obtain the
	original output \citep{BenureauAndRougier2017}. We measured reproducibility
	qualitatively by asking developers if they have any concerns that their
	computational results won't be reproducible and if they have taken any
	steps to ensure reproducibility.

@@

To answer~\rqref{RQ_WhatProjects} we identify existing LBM projects.
We find the candidate software through search engine queries targeting
authoritative lists. LBM software is listed on \href{https://github.com/}
{GitHub} and \href{https://swmath.org/} {swMATH} and found in scholarly
articles. The Domain Expert (Section~\ref{sec_vet_software_list}) helps with
selecting the candidate software.

We consider the following properties when creating the list and reviewing the
candidate software:

@@

The initial list had 45 packages, including some that we later found we could
not use because of incompleteness or an absence of publicly available source
code.

@@

To reduce the length of the list of software packages and to
answer~\rqref{RQ_WhatProjects}, we filter the initial list of 45 packages. We
apply the following filters in the priority order listed.

\begin{enumerate}
	\item Scope: We remove software by narrowing what functionality we
    consider within the scope of the domain.
	\item Usage: We eliminate software if its installation procedure is
	missing or unclear.
	\item Age: We eliminate the older software packages (with age measured by
	the last date of a change), except when an older software package appears to
	be highly recommended and currently in use. 
\end{enumerate}

For the third item in the above filter, we categorized software packages as
`alive' if documentation updates happened within the last 18 months. Packages
were categorized as `dead' if their last update were more than 18 month ago.

@@

Filtering by scope, usage, and age decreased the size of the list to 23
packages. We removed 22 packages for the following reasons: we cannot install
them, they are incomplete, source code is not publicly available, a license is
needed, or the project is out of scope. \citet{Michalski2021} lists the
eliminated software packages. Of the remaining 23 packages, we kept some despite
their `dead` status because of their presence on authoritative LBM software
lists and because of their surface excellence. After the initial measurement
exercise, which took place in mid-2020, we added one more package in January
2022: Musubi. We found Musubi after our initial search, but we felt obliged to
include it since it clearly fits our criteria. Time constraints meant we didn't
update the previous manual data collection, but we did redo the automated data
collection.

The final list of 24 software packages is in two tables.
Table~\ref{alivepackages} lists packages that fell into the `alive' category as
of mid-2020 (January 2022 for Musubi), and Table~\ref{deadpackages} lists
packages that were `dead'. The tables include a hyperlink to each project and,
when available, citations for relevant publications.  The final list of software
packages shows considerable variation in purpose, size, user interfaces, and
programming languages. For example, the OpenLB software package is predominantly
a \CC{} package that uses hybrid parallelization and addresses many CFD problems
\citep{heuveline2009towards}. The software package pyLBM is an all-in-one Python
language package for numerical simulations \citep{graille2017pylbm}. ESPResSo
consists of extensible \CC{} and Python code specifically for research on soft
matter \citep{weik2019espresso}. The HemeLB package simulates fluid flow in
several medical domains and is written predominantly in C, \CC, and Python
\citep{mazzeo2008hemelb}.

@@

The top four packages, ESPREsSo, Ludwig, Palabos, and OpenLB, have all of the
common artifacts, except Palabos does not have theory notes and only three
(ESPREsSo, Ludwig, Palabos) of the packages appear to use a version control
system.  The top four packages also have most of the uncommon artifacts. At
the time of data collection, only one of the four packages (Palabos) did not
have a user manual or guide, but there was a broken link on the package website
indicating that such an artifact might exist. This broken link was later fixed,
but this is not reflected in our data because it was not present at the time of
data collection (mid-2020). Despite the broken link, Palabos does have a
detailed and informative website. As far as we could observe, Ludwig does not
appear to have a publicly visible design and OpenLB does not appear to use a
version control system. However, OpenLB's use of version numbers suggests that
version control may be used privately by the developers.
